---
title: "Lab 2: Probability Distributions for Count Data"
subtitle: "Module 2 - Distributions"
output: html_document
---

## Objectives
- Visualize Poisson and negative binomial distributions
- Fit distributions to RNA-seq data
- Understand overdispersion

## Setup

```{r setup, message=FALSE}
library(ggplot2)
library(MASS)
library(airway)
```

## 1. Poisson Distribution

```{r poisson}
# Poisson: mean = variance
lambda <- 10
x <- 0:30

# Poisson probabilities
poisson_probs <- dpois(x, lambda)

plot(x, poisson_probs, type = "h", lwd = 3, col = "steelblue",
     main = paste("Poisson Distribution (λ =", lambda, ")"),
     xlab = "Count", ylab = "Probability")

# Check mean and variance
cat("Theoretical mean:", lambda, "\n")
cat("Theoretical variance:", lambda, "\n")
```

## 2. Negative Binomial Distribution

```{r negbin}
# Negative binomial: allows variance > mean
mu <- 10
size <- 2  # smaller size = more overdispersion

nb_probs <- dnbinom(x, mu = mu, size = size)

# Compare to Poisson
plot(x, poisson_probs, type = "h", lwd = 3, col = "steelblue",
     main = "Poisson vs Negative Binomial",
     xlab = "Count", ylab = "Probability", ylim = c(0, max(c(poisson_probs, nb_probs))))
lines(x + 0.2, nb_probs, type = "h", lwd = 3, col = "coral")
legend("topright", c("Poisson", "Negative Binomial"), 
       col = c("steelblue", "coral"), lwd = 3)
```

## 3. Effect of Dispersion

```{r dispersion}
par(mfrow = c(1, 3))

sizes <- c(100, 10, 1)  # high to low precision
colors <- c("green", "orange", "red")

for (i in 1:3) {
  probs <- dnbinom(x, mu = 10, size = sizes[i])
  plot(x, probs, type = "h", lwd = 3, col = colors[i],
       main = paste("Size =", sizes[i]),
       xlab = "Count", ylab = "Probability")
}

par(mfrow = c(1, 1))
```

## 4. Fitting to Real Data

```{r fit-real}
data(airway)
counts <- assay(airway)

# Select a highly expressed gene
high_expr_gene <- counts["ENSG00000000003", ]

# Fit Poisson
pois_fit <- fitdistr(high_expr_gene, "Poisson")

# Fit Negative Binomial
nb_fit <- fitdistr(high_expr_gene, "Negative Binomial")

cat("Gene ENSG00000000003\n")
cat("Observed mean:", mean(high_expr_gene), "\n")
cat("Observed variance:", var(high_expr_gene), "\n")
cat("\nPoisson lambda:", pois_fit$estimate, "\n")
cat("NB mu:", nb_fit$estimate["mu"], "\n")
cat("NB size:", nb_fit$estimate["size"], "\n")
```

## 5. Visualize Fit

```{r visualize-fit}
# Histogram of observed data
hist(high_expr_gene, breaks = 20, freq = FALSE, col = "lightgray",
     main = "Observed vs Fitted Distributions",
     xlab = "Count")

# Overlay fitted distributions
x_range <- seq(min(high_expr_gene), max(high_expr_gene), by = 1)

# Poisson
lines(x_range, dpois(x_range, pois_fit$estimate), 
      col = "blue", lwd = 2, type = "b")

# Negative binomial
lines(x_range, dnbinom(x_range, 
                       mu = nb_fit$estimate["mu"], 
                       size = nb_fit$estimate["size"]), 
      col = "red", lwd = 2, type = "b")

legend("topright", c("Observed", "Poisson", "Neg. Binomial"),
       col = c("gray", "blue", "red"), lwd = c(10, 2, 2))
```

## 6. Overdispersion Across Genes

```{r overdispersion}
# Calculate mean and variance for all genes
gene_stats <- data.frame(
  mean = rowMeans(counts),
  var = apply(counts, 1, var)
)
gene_stats <- gene_stats[gene_stats$mean > 0, ]

# Plot
ggplot(gene_stats, aes(x = log10(mean), y = log10(var))) +
  geom_point(alpha = 0.3, size = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", size = 1) +
  labs(x = "Log10(Mean)", y = "Log10(Variance)",
       title = "Overdispersion in RNA-seq Data",
       subtitle = "Red line: Poisson expectation (var = mean)") +
  theme_minimal()
```

## Exercise Questions

1. What happens to the negative binomial shape as `size` decreases?
2. Why does the negative binomial fit better than Poisson?
3. Looking at the overdispersion plot, are most genes over- or under-dispersed?
4. What would the plot look like if Poisson were appropriate?

```{r exercises}
# Your answers here

```

---
*End of Lab 2*

